\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,bm}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc,shapes.geometric}

\title{Road Event Detector Formulations}
\author{road\_events documentation}
\date{\today}

\newcommand{\dt}{\Delta t}
\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\sgn}{\operatorname{sgn}}

\begin{document}
\maketitle
\tableofcontents
\newpage

\section{Scope}

This document records the math and streaming algorithms implemented by the
\path{road_events} crate. The crate detects road events from vehicle-frame
motion signals produced by the simulation and visualization pipeline. The
detectors are intentionally small-state and embedded-friendly: they process one
sample at a time, use bounded memory, and avoid FFTs, dynamic allocation, and
offline window buffering.

The detectors are not part of the EKF state estimator. They consume already
estimated or referenced motion channels such as vehicle speed, vehicle pitch,
gravity-compensated vertical acceleration, and yaw rate. The current event set
is:
\begin{itemize}[nosep]
  \item speed bumps,
  \item sustained uphill and downhill intervals,
  \item reverse-driving intervals,
  \item harsh acceleration,
  \item harsh braking,
  \item harsh cornering.
\end{itemize}

\section{Signals and Units}

\begin{longtable}{>{\raggedright\arraybackslash}p{0.22\linewidth} >{\raggedright\arraybackslash}p{0.20\linewidth} >{\raggedright\arraybackslash}p{0.48\linewidth}}
\toprule
\textbf{Symbol} & \textbf{Unit} & \textbf{Meaning} \\
\midrule
\(t_k\) & s & Sample timestamp. \\
\(v_x\) & m/s & Vehicle-frame forward velocity. Positive is forward; negative is reverse. \\
\(s=\abs{v_x}\) & m/s & Speed magnitude used by threshold logic. \\
\(\theta\) & deg & Vehicle pitch angle. Positive pitch is uphill in the current detector convention. \\
\(a_z\) & m/s\(^2\) & Gravity-compensated vehicle-frame vertical acceleration. \\
\(\omega_z\) & rad/s & Vehicle yaw rate. \\
\bottomrule
\end{longtable}

The speed-bump detector uses pitch and vertical acceleration. Hill detection
uses pitch and speed. Reverse and harsh longitudinal detection use forward
velocity. Harsh cornering uses speed and yaw rate.

\begin{figure}[h]
\centering
\vspace{0.4cm}
\begin{tikzpicture}[
  node distance=0.85cm and 1.0cm,
  src/.style={draw,rounded corners,fill=blue!8,align=center,minimum width=2.9cm,minimum height=0.75cm},
  det/.style={draw,rounded corners,fill=green!8,align=center,minimum width=3.0cm,minimum height=0.75cm},
  eventnode/.style={draw,rounded corners,fill=orange!10,align=center,minimum width=2.5cm,minimum height=0.75cm},
  arr/.style={-{Latex[length=2mm]},thick}
]
  \node[src] (motion) {vehicle-frame\\motion signals};
  \node[det,right=of motion] (detectors) {streaming\\event detectors};
  \node[eventnode,right=of detectors] (events) {map/trace\\event intervals};
  \draw[arr] (motion) -- node[above=7mm,fill=white,inner sep=1pt]{one sample} (detectors);
  \draw[arr] (detectors) -- node[above=7mm,fill=white,inner sep=1pt]{optional event} (events);
\end{tikzpicture}
\caption{All event detectors run as streaming post-processors over vehicle
motion signals.}
\end{figure}

\section{Shared Streaming Primitives}

\subsection{Timestamp Handling}

The implementation clamps per-sample time deltas to keep bad or repeated
timestamps from creating unbounded filter updates:
\begin{equation}
  \dt_k = \operatorname{clamp}(t_k-t_{k-1}, 0, 0.2).
\end{equation}
For high-pass and derivative paths that require a strictly positive time step,
the lower bound is \(10^{-3}\) or \(10^{-4}\) seconds depending on the path.

\subsection{First-Order High-Pass Filter}

The speed-bump detector removes road grade and slow body motion with the
standard discrete first-order high-pass filter:
\begin{align}
  \tau &= \frac{1}{2\pi f_c},\\
  \alpha_k &= \frac{\tau}{\tau+\dt_k},\\
  y_k &= \alpha_k(y_{k-1}+u_k-u_{k-1}).
\end{align}
This is applied independently to vehicle pitch and gravity-compensated vertical
acceleration.

\subsection{Exponential Moving Average}

The crate uses an exponential moving average (EMA) for adaptive noise floors
and smoothed velocity derivatives:
\begin{align}
  \alpha_k &= \frac{\dt_k}{\max(\tau,\dt_k)+\dt_k},\\
  \bar{x}_k &= (1-\alpha_k)\bar{x}_{k-1}+\alpha_k x_k .
\end{align}
For noise floors, the input is the absolute high-pass value:
\begin{equation}
  \sigma_k = \operatorname{EMA}(\sigma_{k-1}, \abs{y_k}, \tau, \dt_k).
\end{equation}
The EMA is a low-memory substitute for a sliding RMS window. It provides slow
adaptation to background vibration without storing historical samples.

\subsection{Interval Accumulation}

Most detectors convert an instantaneous metric \(m_k\) into an event interval
using enter and exit thresholds:
\begin{align}
  \text{enter if}\quad &m_k \ge m_{\mathrm{enter}},\\
  \text{stay active while}\quad &m_k \ge m_{\mathrm{exit}},
  \qquad m_{\mathrm{exit}} < m_{\mathrm{enter}}.
\end{align}
During an active interval the detector accumulates:
\begin{align}
  T &= \sum_k \dt_k,\\
  \bar{m} &= \frac{1}{T}\sum_k m_k\dt_k,\\
  m_{\max} &= \max_k m_k,\\
  \bar{s} &= \frac{1}{T}\sum_k s_k\dt_k.
\end{align}
The interval is emitted only if \(T\ge T_{\min}\). A refractory time suppresses
duplicate events immediately after a valid event.

\section{Speed-Bump Detector}

\subsection{Physical Signature}

A road speed bump creates a short vertical impulse sequence as the front axle
and rear axle traverse the bump. In gravity-compensated vehicle-frame vertical
acceleration, the idealized pattern is an alternating sequence of three extrema:
\begin{equation}
  a_{z,\mathrm{hp}}(t_1),\quad a_{z,\mathrm{hp}}(t_2),\quad a_{z,\mathrm{hp}}(t_3),
  \qquad
  \sgn(a_1)=\sgn(a_3)\ne\sgn(a_2).
\end{equation}
Pitch motion is used as corroboration rather than as the primary trigger,
because pitch angle alone can also respond to braking, turning on uneven roads,
or grade changes.

\subsection{Filtered Signals}

The detector computes:
\begin{align}
  \theta_{\mathrm{hp},k} &= \operatorname{HPF}(\theta_k; f_c=0.45~\mathrm{Hz}),\\
  a_{\mathrm{hp},k} &= \operatorname{HPF}(a_{z,k}; f_c=0.70~\mathrm{Hz}).
\end{align}
Adaptive noise floors are:
\begin{align}
  \sigma_{\theta,k} &= \operatorname{EMA}(\abs{\theta_{\mathrm{hp},k}};\tau=6~\mathrm{s}),\\
  \sigma_{a,k} &= \operatorname{EMA}(\abs{a_{\mathrm{hp},k}};\tau=6~\mathrm{s}).
\end{align}
The peak thresholds are:
\begin{align}
  A_{\mathrm{thr}} &= \max(3.5\,\sigma_a,\;1.5~\mathrm{m/s^2}),\\
  \Theta_{\mathrm{thr}} &= \max(3.0\,\sigma_\theta,\;0.25^\circ).
\end{align}

\subsection{Extremum Extraction}

The vertical acceleration slope is
\begin{equation}
  \dot{a}_{\mathrm{hp},k}
  = \frac{a_{\mathrm{hp},k}-a_{\mathrm{hp},k-1}}{\dt_k}.
\end{equation}
A local maximum is detected when \(\dot{a}_{k-1}>0\) and \(\dot{a}_{k}\le 0\);
a local minimum is detected when \(\dot{a}_{k-1}<0\) and \(\dot{a}_{k}\ge 0\).
An extremum is kept only when its absolute vertical acceleration exceeds the
physical floor \(1.5~\mathrm{m/s^2}\). The detector stores only the latest four
extrema, so memory use remains constant.

For each extremum interval, the detector also stores the maximum pitch
high-pass magnitude since the previous extremum:
\begin{equation}
  \Theta_i = \max_{t\in(t_{i-1},t_i]}\abs{\theta_{\mathrm{hp}}(t)}.
\end{equation}

\subsection{Candidate Timing}

Each latest-three-extrema group is scored only if the polarities alternate. Let
\begin{equation}
  D = t_3-t_1,\qquad
  \bar{s}=\frac{s_1+s_2+s_3}{3}.
\end{equation}
The candidate must pass both absolute and speed-adaptive duration bounds:
\begin{align}
  D_{\min} &=
  \max\left(0.18,\;0.35\frac{L_{\min}}{\bar{s}}\right),\\
  D_{\max} &=
  \min\left(1.8,\;2.5\frac{L_{\max}}{\bar{s}}\right),
\end{align}
where \(L_{\min}=1.8~\mathrm{m}\) and \(L_{\max}=3.6~\mathrm{m}\) are plausible
front/rear axle spacing bounds. The speed must also satisfy
\(\bar{s}\ge1.5~\mathrm{m/s}\).

The detector tracks the total time within the pattern for which
\(\abs{a_{\mathrm{hp}}}\ge1.5~\mathrm{m/s^2}\). It requires:
\begin{align}
  T_{\mathrm{active}} &\ge 0.25~\mathrm{s},\\
  \frac{T_{\mathrm{active}}}{D} &\ge 0.25.
\end{align}

\subsection{Score and Confidence}

For extrema magnitudes \(A_i=\abs{a_i}\), pitch peaks \(\Theta_i\), and
candidate duration \(D\), the component scores are:
\begin{align}
  A_{\max} &= \max_i A_i,\\
  \Theta_{\max} &= \max_i \Theta_i,\\
  S_a &= \operatorname{clamp}\left(\frac{A_{\max}}{A_{\mathrm{thr}}}-1,\;0,\;1\right),\\
  S_\theta &= \operatorname{clamp}\left(\frac{\Theta_{\max}}{\Theta_{\mathrm{thr}}}-1,\;0,\;1\right),\\
  S_b &= \frac{\min(A_1,A_3)}{\max(A_1,A_3,10^{-3})}.
\end{align}
The spacing score favors the center of the admissible duration interval:
\begin{align}
  D_c &= \frac{D_{\min}+D_{\max}}{2},&
  W &= \frac{\max(D_{\max}-D_{\min},10^{-3})}{2},\\
  S_D &= \operatorname{clamp}\left(1-\frac{\abs{D-D_c}}{W},\;0,\;1\right).
\end{align}
The final pattern score is:
\begin{equation}
  S =
  \operatorname{clamp}\left(
    (0.45S_a+0.25S_D+0.15S_b+0.15S_\theta)S_\theta,\;0,\;1
  \right).
\end{equation}
The pitch score multiplies the weighted score, so vertical acceleration alone
does not emit an event.

An event is emitted when \(S\ge0.12\) and at least four seconds have elapsed
since the last bump event. The UI-facing confidence is not a calibrated
probability; it is normalized only after the trigger threshold is cleared:
\begin{equation}
  c_{\mathrm{display}} =
  0.90 + 0.08\,
  \operatorname{clamp}\left(\frac{S-0.12}{1-0.12},0,1\right).
\end{equation}

\section{Hill Detector}

The hill detector identifies sustained grade from vehicle pitch. It is a
piecewise-constant interval detector with sign:
\begin{equation}
  \mathrm{kind}(\theta)=
  \begin{cases}
    \mathrm{uphill}, & \theta \ge 4^\circ,\\
    \mathrm{downhill}, & \theta \le -4^\circ,\\
    \varnothing, & \text{otherwise.}
  \end{cases}
\end{equation}
An interval is emitted when the same signed condition remains active for at
least three seconds. During the interval, the detector reports:
\begin{align}
  T &= t_{\mathrm{end}}-t_{\mathrm{start}},\\
  \bar{\theta} &= \frac{1}{T}\int_{t_{\mathrm{start}}}^{t_{\mathrm{end}}}\theta(t)\,dt,\\
  \theta_{\max} &= \max_t \abs{\theta(t)},\\
  \bar{s} &= \frac{1}{T}\int_{t_{\mathrm{start}}}^{t_{\mathrm{end}}} \max(s(t),0)\,dt.
\end{align}
The implementation emits the event when pitch exits the threshold band or when
the detector is explicitly finished at the end of a replay.

\section{Reverse Detector}

Reverse detection uses vehicle-frame forward velocity directly. The detector
starts a candidate when:
\begin{equation}
  v_x < -0.5~\mathrm{m/s}.
\end{equation}
The candidate is confirmed only after this condition persists for
\(0.5~\mathrm{s}\). Once confirmed, the interval remains active while:
\begin{equation}
  v_x < -0.2~\mathrm{m/s}.
\end{equation}
This hysteresis prevents repeated short events around zero speed. The event is
closed after \(0.5~\mathrm{s}\) above the exit threshold and emitted if its
duration is at least one second.

Reported reverse speed uses positive magnitude:
\begin{equation}
  s_{\mathrm{rev}}=\max(-v_x,0).
\end{equation}
Mean and peak reverse speeds are accumulated over the confirmed interval.

\section{Harsh Acceleration and Braking}

\subsection{Velocity-Derivative Metric}

Harsh longitudinal events are based on changes in forward velocity rather than
raw accelerometer samples. This makes the detector less sensitive to road
vibration and IMU mounting artifacts. The raw derivative is:
\begin{equation}
  a_{x,\mathrm{raw},k} =
  \operatorname{clamp}\left(
    \frac{v_{x,k}-v_{x,k-1}}{\dt_k},\;-15,\;15
  \right)~\mathrm{m/s^2}.
\end{equation}
The metric used by the interval detector is an EMA-smoothed derivative:
\begin{equation}
  \hat{a}_{x,k} = \operatorname{EMA}(\hat{a}_{x,k-1},a_{x,\mathrm{raw},k},\tau=0.6~\mathrm{s},\dt_k).
\end{equation}
The speed gate uses:
\begin{equation}
  s_k=\max(\abs{v_{x,k}},\abs{v_{x,k-1}}).
\end{equation}

\subsection{Harsh Acceleration}

If \(s_k<1~\mathrm{m/s}\), the metric is forced to zero. Otherwise:
\begin{equation}
  m_{\mathrm{accel},k}=\max(\hat{a}_{x,k},0).
\end{equation}
The interval thresholds are:
\begin{align}
  m_{\mathrm{enter}} &= 2.5~\mathrm{m/s^2},\\
  m_{\mathrm{exit}} &= 2.0~\mathrm{m/s^2},\\
  T_{\min} &= 0.4~\mathrm{s},\\
  T_{\mathrm{refractory}} &= 2.0~\mathrm{s}.
\end{align}
The emitted event reports duration, mean and peak acceleration metric, mean and
peak speed, and net velocity change:
\begin{equation}
  \Delta v_x = v_{x,\mathrm{end}}-v_{x,\mathrm{start}}.
\end{equation}

\subsection{Harsh Braking}

Braking uses the same smoothed derivative but flips the sign:
\begin{equation}
  m_{\mathrm{brake},k}=\max(-\hat{a}_{x,k},0).
\end{equation}
The default thresholds are:
\begin{align}
  m_{\mathrm{enter}} &= 3.0~\mathrm{m/s^2},\\
  m_{\mathrm{exit}} &= 2.4~\mathrm{m/s^2},\\
  T_{\min} &= 0.4~\mathrm{s},\\
  T_{\mathrm{refractory}} &= 2.0~\mathrm{s}.
\end{align}
The reported \(\Delta v_x\) is signed, so braking events normally have
\(\Delta v_x<0\).

\section{Harsh Cornering}

Harsh cornering uses the steady-turn kinematic relation:
\begin{equation}
  a_y \approx v\omega_z .
\end{equation}
The detector metric is:
\begin{equation}
  m_{\mathrm{corner},k} =
  \begin{cases}
    \abs{s_k\omega_{z,k}}, & s_k\ge3~\mathrm{m/s},\\
    0, & s_k<3~\mathrm{m/s}.
  \end{cases}
\end{equation}
The default interval thresholds are:
\begin{align}
  m_{\mathrm{enter}} &= 3.0~\mathrm{m/s^2},\\
  m_{\mathrm{exit}} &= 2.4~\mathrm{m/s^2},\\
  T_{\min} &= 0.5~\mathrm{s},\\
  T_{\mathrm{refractory}} &= 2.0~\mathrm{s}.
\end{align}
The steady-turn approximation is deliberately simple. It is robust for
embedded event detection because it only needs speed and yaw rate, and it does
not depend on lateral accelerometer calibration or road bank compensation.

\section{Default Configuration Summary}

\begin{longtable}{>{\raggedright\arraybackslash}p{0.30\linewidth} >{\raggedright\arraybackslash}p{0.24\linewidth} >{\raggedright\arraybackslash}p{0.36\linewidth}}
\toprule
\textbf{Detector} & \textbf{Default} & \textbf{Purpose} \\
\midrule
Speed bump pitch HPF & \(0.45~\mathrm{Hz}\) & Reject grade and slow pitch. \\
Speed bump vertical accel HPF & \(0.70~\mathrm{Hz}\) & Reject slow body motion. \\
Speed bump noise EMA & \(6.0~\mathrm{s}\) & Adaptive vibration floor. \\
Speed bump minimum speed & \(1.5~\mathrm{m/s}\) & Avoid static/creep false positives. \\
Speed bump duration & \(0.18\)--\(1.8~\mathrm{s}\) & Absolute event timing. \\
Speed bump wheelbase bounds & \(1.8\)--\(3.6~\mathrm{m}\) & Speed-adaptive timing. \\
Hill pitch threshold & \(4^\circ\) & Sustained uphill/downhill grade. \\
Hill minimum duration & \(3.0~\mathrm{s}\) & Suppress short pitch transients. \\
Reverse enter/exit & \(-0.5/-0.2~\mathrm{m/s}\) & Hysteresis around zero speed. \\
Reverse minimum duration & \(1.0~\mathrm{s}\) & Suppress brief backing corrections. \\
Harsh accel enter/exit & \(2.5/2.0~\mathrm{m/s^2}\) & Longitudinal acceleration hysteresis. \\
Harsh brake enter/exit & \(3.0/2.4~\mathrm{m/s^2}\) & Longitudinal deceleration hysteresis. \\
Harsh corner enter/exit & \(3.0/2.4~\mathrm{m/s^2}\) & Lateral acceleration hysteresis. \\
\bottomrule
\end{longtable}

\section{Visualizer Interpretation}

The map markers and hover mini-plots are visualization aids. They do not feed
back into the detectors. For each event, the visualizer draws a short local
window around the detected interval, currently with five seconds of padding on
both sides when data is available:
\begin{equation}
  [t_{\mathrm{start}}-5,\;t_{\mathrm{end}}+5].
\end{equation}
The mini-plot shows the signal that caused the detector to trigger, for example
vertical acceleration and pitch for speed bumps, longitudinal acceleration for
harsh acceleration/braking, and yaw-rate-derived lateral acceleration for harsh
cornering.

\section{Design Rationale}

The event crate favors deterministic low-memory logic over more expressive
offline methods. Windowed FFTs, matched filters, and machine-learning
classifiers can be useful for richer offline analysis, but they introduce
larger buffers, more tuning surface, and more deployment complexity. The current
detectors instead use physically interpretable metrics:
\begin{itemize}[nosep]
  \item speed bumps: alternating high-pass vertical acceleration extrema
    corroborated by pitch motion,
  \item hills: sustained pitch sign and magnitude,
  \item reverse: signed vehicle-frame forward velocity with hysteresis,
  \item harsh acceleration/braking: smoothed velocity derivative,
  \item harsh cornering: steady-turn lateral acceleration \(v\omega_z\).
\end{itemize}

This keeps the algorithms easy to port to small embedded targets and easy to
debug from the visualizer.

\end{document}
